Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-07 paper.Rnw
some of aristotle's comments addressed

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5]{Aristotle Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such a cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example \citet{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations; or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

@@

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
\citet{Yushkevich2009} acquired and labelled hippocampal subfields on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, they demonstrate accurate parcellation of hippocampal subfields in
unlabelled MRI volumes.

Here we address the issue of limiting the number of input atlases by tuning our
algorithm, for segmentation of the entire hippocampus, using a multi-fold
experiment performed on a subset Alzheimer's Disease Neuroimaging Initiative
(ADNI) 1 dataset. Based on the parameters we find in our experiment, we validate
our algorithm using all of the data available in the ADNI Complete 1Yr sample
and compare our segmentations to the other segmentations that are available
through the ADNI informatics portal.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also apply our
segmentations to a dataset of normal controls and individuals suffering from
first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image.  {\em Label propagation}, or {\em labelling}, is the 
process by which two images are registered and the resulting transformation 
is applied to the labels from one image to bring them into alignment with the 
other image. We will use the term {\em atlas} to mean a manually segmented image,
and the term {\em template} to mean an automatically segmented image (i.e. 
via label propagation). The terms {\em atlas library} and {\em template library} 
describe any set of such images.Additionally, we will use the term {\em target} 
to refer to an unlabelled image that is undergoing segmentation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)
caption = paste(
  "\\textbf{ADNI-1 cross-validation subset demographics.}", 
  "Clinical Dementia Rating-Sum of Boxes (CDR-SB), Alzheimer's Disease",
  "Assessment Scale (ADAS), Mini–Mental State Examination (MMSE).", sep=" ") 
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in \citet{Robbins2004}, displayed in Table
\ref{tab:ANIMAL-params}.

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as
the gold standard to compare with \citep{Hsu2002}.  The segmentation accuracy
reported is averaged over the ten validation rounds for each parameter setting.

In order to investigate the performance of \mb in a real world setting in which 
only one set of atlas and template images are used, we explore the variability 
in label agreement at fixed parameter settings when the choice for atlas and 
template images is varied. This is achieved by first computing the standard 
deviation and variance of DSC scores in each block of ten validation rounds per
subject. The distribution of these statistics across all subjects is then
compared between \mb and multi-atlas using a Student's t-test. A significant
difference between distributions is taken to show either a larger or smaller
level of variability between methods.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Subfield Validation Template library demographics (healthy controls)
%                                     
<<WAval-healthy-demographics, echo=F, results="asis",cache=FALSE>>=
master = read.csv('data/DT_study.csv')
tab <- summary(~Age + Sex + Education + Handedness, 
               data = master, method="reverse", na.rm = TRUE, test=FALSE)
caption = paste(
  "\\textbf{ADNI-1 cross-validation subset demographics.}", 
  "Clinical Dementia Rating-Sum of Boxes (CDR-SB), Alzheimer's Disease",
  "Assessment Scale (ADAS), Mini–Mental State Examination (MMSE).", sep=" ") 
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:WAval-healthy-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 3: Application to the segmentation first episode
schizophrenia patients}

@@

\paragraph{\FEPdataset dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated.  This poses a problem for
evaluation by measuring overlap.  That is, since different protocols will
necessarily produce segmentations that do not perfectly overlap, the degree of
overlap cannot be solely used to compare segmentation methods using different
protocols. In place of an overlap metric, we can assess the degree of correlation in
average hippocampal volume of the subjects produced by each method.
Specifically, Pearson correlation is used.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First Episode SZ Demographics Table
%
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
caption = paste(
  "\\textbf{Schizophrenia First Episode Patient Demographics.}",
  "Clinical Dementia Rating-Sum of Boxes (CDR-SB), Alzheimer's Disease",
  "State Examination (MMSE).", sep=" ") 
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:SZFEP-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{ADNI1 dataset revisited} 
All images from the {\em \adnidataset} standardized dataset described
in Experiment 1 are used.  Clinical and demographic data are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the \adnidataset dataset are
segmented. The optimal size of template library, registration method, and label
fusion method found in Experiment 1 are used. The template library is composed
of equal numbers of images from each disease class (AD, MCI, and cognitively
normal controls).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
<<ADNI1-Dataset-Demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]
yr1 = 
adnimerge.yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge.yr1, 
               #subset = RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)
caption = paste(
  "\\textbf{ADNI1 1.5T Complete 1Yr dataset demographics.}", 
  "Clinical Dementia Rating-Sum of Boxes (CDR-SB), Alzheimer's Disease",
  "Assessment Scale (ADAS), Mini–Mental State Examination (MMSE).", sep=" ") 
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
    caption=caption, caption.loc = c("top"),
    title="tab:ADNI1-Dataset-Demographics",
    label="tab:ADNI1-Dataset-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,7]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5,7]{Aristotle N. Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[7]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater.
However with the availability of increasingly large MRI datasets, the time and
expertise required for manual segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}.  Further, there is
little agreement between researchers regarding how exactly the hippocampus
should be identified in MRI images \citep{Geuze2004} and this has led to efforts
to create an unified hippocampal segmentation protocol \citep{Jack2011}.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example \citet{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations, or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}).  Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images.  Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

@@

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, it  demonstrates that the
applicability of hippocampal subfield segmentation using data from 10 subjects.
In the second method, \citet{Yushkevich2009} acquired and labelled hippocampal
subfields on high-resolution MRI data from post-mortem medial temporal lobe
samples. Using nonlinear registration guided by manually derived hippocampus
masks and specific landmarks, they demonstrate accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

In this paper we undertake a thorough validation of the \mb algorithm for
segmentation of the hippocampus.  First, we address the feasibility and accuracy
of whole hippocampus segmentation with a limited number of input atlases by
performing a multi-fold experiment using a subset of the Alzheimer's Disease
Neuroimaging Initiative (ADNI) 1 dataset. Based on the parameters we find in our
experiment, we then validate our algorithm using all of the data available in
the ADNI Complete 1Yr sample and compare our segmentations to the other
segmentations that are available through the ADNI informatics portal. To ensure
that we have not over-fit our parameters to the aging or neurodegenerative
brain, we also apply our segmentations to a dataset of individuals suffering
from first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image.  {\em Label propagation}, or {\em labelling}, is the 
process by which two images are registered and the resulting transformation 
is applied to the labels from one image to bring them into alignment with the 
other image. We will use the term {\em atlas} to mean a manually segmented image,
and the term {\em template} to mean an automatically segmented image (i.e. 
via label propagation). The terms {\em atlas library} and {\em template library} 
describe any set of such images. Additionally, we will use the term {\em target} 
to refer to an unlabelled image that is undergoing segmentation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as
the gold standard with which to compare \citep{Hsu2002}. The segmentation
accuracy reported is averaged over the ten validation rounds for each parameter
setting.

In order to investigate the performance of \mb in a real world setting in which 
only one size of atlas and template libraries are used, we explore the
variability in label agreement at fixed parameter settings under varying  
atlas and template library compositions. This is achieved by first computing the
standard deviation and variance of DSC scores in each block of ten validation
rounds per subject. The distribution of these statistics across all subjects is
then compared between \mb and multi-atlas using a Student's t-test. A
significant difference between distributions is taken to show either a larger or
smaller level of variability between methods.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Subfield Validation Template library demographics (healthy controls)
%                                     
<<WAval-healthy-demographics, echo=F, results="asis",cache=FALSE>>=
master = read.csv('data/DT_study.csv')
tab <- summary(~Age + Sex + Education + Handedness, 
               data = master, method="reverse", na.rm = TRUE, test=FALSE)
caption = 
  "\\textbf{Subfield validation healthy control demographics.}
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:WAval-healthy-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 3: Application to the segmentation of first episode
schizophrenia patients}

@@

\paragraph{\FEPdataset dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15 to 30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated. This difference poses a
problem for evaluation by measuring overlap.  That is, since different protocols
will necessarily produce segmentations that do not perfectly overlap, the degree
of overlap cannot be solely used to compare segmentation methods using different
protocols. In place of an overlap metric, we can assess the degree of
correlation in average hippocampal volume of the subjects produced by each
method. Specifically, a Pearson correlation is used.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First Episode SZ Demographics Table
%
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
caption = 
  "\\textbf{Schizophrenia First Episode Patient Demographics.}
  ambi - ambidextrous. 
  SES  - Socioeconomic Status score. 
  FSIQ - Full Scale IQ.
"
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:SZFEP-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{ADNI1 dataset revisited} 
All images $1919$ images from the {\em \adnidataset} standardized dataset
described in Experiment 1 are used.  Clinical and demographic data for these
subjects are shown in Table \ref{tab:ADNI1-Dataset-Demographics}.

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and used majority vote label fusion.
The entire \adnidataset dataset was segmented by \mb and we now compare the
resulting volumes with those obtained by manual segmentation (SNT), and other
automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection. Quality control for FreeSurfer segmentations was provided along
with the segmentations from ADNI \cite{Hartig2010}. One of the authors (MP) 
performed visual quality inspection for MAGeT and FSL segmentations using similar 
quality control guidelines (if either hippocampus was under or over segmented
by greater than 10mm in three or more slices then the segmentation did not
pass).  Only those images which had quality segmentations from each method are
included in the following analysis (a total of
\Sexpr{length(means.complete$RID)} images).

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the \adnidataset dataset are segmented. The
optimal size of template library, registration method, and label fusion method
found in Experiment 1 are used. The template library is composed of 21 randomly
chosen subject images (7 from each disease class).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
<<ADNI1-Dataset-Demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]
yr1 = 
adnimerge.yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge.yr1, 
               #subset = RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI1 1.5T Complete 1Yr dataset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
    caption=caption, caption.loc = c("top"),
    title="tab:ADNI1-Dataset-Demographics",
    label="tab:ADNI1-Dataset-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
